---
title: "Analysis of Covariance (ANCOVA)"
---

## What a reviewer wants

ANCOVA is the workhorse for a continuous endpoint (e.g. change from baseline)
adjusted for the baseline value and, often, stratification factors. The quantity
of interest is the **adjusted treatment difference** (active minus placebo LS
mean) with its confidence interval. The minimal, common-package way to get this
is base R's `lm()` for the fit and **`emmeans`** for the least-squares means —
`emmeans` is the de-facto standard for adjusted means in the R-in-pharma world.

## Minimal analysis

```{r}
#| label: minimal-ancova
#| message: false
library(emmeans)

# Example ADaM-like data: change from baseline, adjusting for baseline.
set.seed(42)
n <- 100
adam <- data.frame(
  USUBJID = seq_len(2 * n),
  TRT01P  = factor(rep(c("Placebo", "Active"), each = n),
                   levels = c("Placebo", "Active")),
  BASE    = rnorm(2 * n, mean = 50, sd = 8)
)
# True model: 4-point active benefit, baseline slope of 0.6.
adam$CHG <- 2 + 0.6 * (adam$BASE - 50) +
  ifelse(adam$TRT01P == "Active", 4, 0) + rnorm(2 * n, sd = 5)

# The entire analysis is two lines.
fit <- lm(CHG ~ TRT01P + BASE, data = adam)
emmeans(fit, ~ TRT01P) |>
  contrast("revpairwise", infer = TRUE)   # Active - Placebo, with CI and p-value
```

`lm()` fits the ANCOVA; `emmeans()` produces the LS means at the mean baseline;
`contrast("revpairwise", ...)` gives the adjusted **Active − Placebo**
difference with its 95% CI. That is the number that goes in the results table,
produced by code a reviewer can read in seconds.

::: {.callout-note}
## Why `emmeans` and not hand-rolled LS means
You *can* compute the adjusted difference from `coef(fit)` directly (for this
simple model it is just the `TRT01P` coefficient). `emmeans` earns its place the
moment you add a stratification factor or an interaction, because it makes the
marginal quantity — and the reference grid it is averaged over — explicit and
auditable. For the simplest model, both agree exactly; the simulation below
uses the raw coefficient to show the estimator itself is sound.
:::

## Simulation evidence: does ANCOVA recover the true effect?

We simulate trials from a data-generating process with a **known treatment
effect of 4**, fit the minimal ANCOVA, and record the estimated treatment
difference, its SE, and its 95% CI. A valid analysis should show ~zero bias and
~95% coverage. We also confirm ANCOVA's efficiency gain over an unadjusted
two-sample comparison.

```{r}
#| label: sim-setup
source("R/sim_utils.R")  # project root is the working dir (see _quarto.yml)

TRUE_EFFECT <- 4
BASE_SLOPE  <- 0.6
N_PER_ARM   <- 100
N_REP       <- 2000

gen <- function(i) {
  base <- rnorm(2 * N_PER_ARM, mean = 50, sd = 8)
  trt  <- factor(rep(c("Placebo", "Active"), each = N_PER_ARM),
                 levels = c("Placebo", "Active"))
  chg  <- 2 + BASE_SLOPE * (base - 50) +
    ifelse(trt == "Active", TRUE_EFFECT, 0) + rnorm(2 * N_PER_ARM, sd = 5)
  data.frame(CHG = chg, TRT01P = trt, BASE = base)
}

# Adjusted analysis: ANCOVA.
fit_ancova <- function(d) {
  m  <- lm(CHG ~ TRT01P + BASE, data = d)
  co <- summary(m)$coefficients["TRT01PActive", ]
  ci <- confint(m)["TRT01PActive", ]
  list(estimate = co["Estimate"], se = co["Std. Error"],
       lower = ci[1], upper = ci[2])
}

# Unadjusted analysis: two-sample comparison ignoring baseline.
fit_unadj <- function(d) {
  m  <- lm(CHG ~ TRT01P, data = d)
  co <- summary(m)$coefficients["TRT01PActive", ]
  ci <- confint(m)["TRT01PActive", ]
  list(estimate = co["Estimate"], se = co["Std. Error"],
       lower = ci[1], upper = ci[2])
}
```

```{r}
#| label: sim-run
res_ancova <- run_simulation(N_REP, gen, fit_ancova, seed = 300)
res_unadj  <- run_simulation(N_REP, gen, fit_unadj,  seed = 300)

comparison <- rbind(
  cbind(method = "ANCOVA (adjusted)",   evaluate_estimator(
    res_ancova$estimate, res_ancova$se, TRUE_EFFECT,
    res_ancova$lower, res_ancova$upper)),
  cbind(method = "Unadjusted",          evaluate_estimator(
    res_unadj$estimate, res_unadj$se, TRUE_EFFECT,
    res_unadj$lower, res_unadj$upper))
)
print_eval(comparison)
```

What to read from the table:

- **Both** estimators are unbiased for the true effect of 4 (baseline is
  balanced by randomisation), and **both** cover at ~95%.
- **ANCOVA's `emp_se` and `rmse` are meaningfully smaller.** Adjusting for the
  prognostic baseline removes variance, so the adjusted estimate is closer to
  the truth on average. That efficiency — not bias removal — is why ANCOVA is
  the standard, and the simulation quantifies exactly how much is gained.
- `se_ratio` near 1 in both rows confirms the reported SEs are honest.

## Takeaways

- The submission-ready ANCOVA is `lm(CHG ~ TRT + BASE)` + `emmeans` for the
  adjusted difference — minimal and reviewer-legible.
- Simulation confirms the estimator is unbiased with correct coverage, and
  quantifies the precision gain over an unadjusted analysis.

```{r}
#| label: session
sessionInfo()
```
