---
title: "Summary Statistics"
---

## What a reviewer wants

For a continuous endpoint, the descriptive table (n, mean, standard deviation,
median, minimum, maximum, and the count of missing values) per treatment arm is
the most-read table in the submission. It should be produced by code a reviewer
can read in one pass. **Base R is enough** — no package is required, so there is
nothing to trust beyond the R interpreter itself.

## Minimal analysis

We build a tiny, explicit summary function and apply it by treatment arm. Every
line maps directly to a number a reviewer will look for.

```{r}
#| label: minimal-summary

# A small, self-documenting summary of one numeric vector.
descr <- function(x) {
  x_obs <- x[!is.na(x)]
  c(
    n       = length(x_obs),
    missing = sum(is.na(x)),
    mean    = mean(x_obs),
    sd      = sd(x_obs),
    median  = median(x_obs),
    min     = min(x_obs),
    max     = max(x_obs)
  )
}

# Example data: change from baseline by treatment arm.
set.seed(1)
adam <- data.frame(
  TRT01P = rep(c("Placebo", "Active"), each = 60),
  CHG    = c(rnorm(60, mean = 1.0, sd = 4),
             rnorm(60, mean = 3.5, sd = 4))
)
adam$CHG[c(5, 20, 95)] <- NA  # a few dropouts

# One line produces the by-arm summary table.
summary_table <- t(sapply(split(adam$CHG, adam$TRT01P), descr))
round(summary_table, 2)
```

That is the whole analysis: a function every reviewer can verify by eye, and a
single `split()` + `sapply()` to apply it by arm.

## Simulation evidence: do the estimates recover the truth?

A summary statistic is only useful if it is an unbiased, well-calibrated
estimate of the population quantity. We simulate many trials from a **known**
mean and standard deviation and check that:

- the sample **mean** is unbiased for the true mean, and its 95% CI covers the
  truth ~95% of the time;
- the sample **standard deviation** is (nearly) unbiased for the true SD.

```{r}
#| label: sim-setup
source("R/sim_utils.R")  # project root is the working dir (see _quarto.yml)

TRUE_MEAN <- 3.5
TRUE_SD   <- 4.0
N_PER_ARM <- 60
N_REP     <- 2000

# Generate one arm's worth of data.
gen <- function(i) {
  data.frame(y = rnorm(N_PER_ARM, mean = TRUE_MEAN, sd = TRUE_SD))
}

# Fit = compute the sample mean and its standard error (one-sample t interval).
fit_mean <- function(d) {
  m  <- mean(d$y)
  se <- sd(d$y) / sqrt(nrow(d))
  tcrit <- qt(0.975, df = nrow(d) - 1)
  list(estimate = m, se = se,
       lower = m - tcrit * se, upper = m + tcrit * se)
}
```

```{r}
#| label: sim-run
res_mean <- run_simulation(N_REP, gen, fit_mean, seed = 100)
print_eval(evaluate_estimator(res_mean$estimate, res_mean$se,
                              truth = TRUE_MEAN,
                              lower = res_mean$lower, upper = res_mean$upper))
```

Read the row as: **bias** near 0, **se_ratio** near 1 (the reported SE matches
the true spread of estimates), and **coverage** near 0.95. That is exactly what
an unbiased, correctly-calibrated estimator should show.

We can check the standard deviation the same way. The sample SD is known to be
slightly *biased low* in small samples; the simulation makes that visible rather
than assumed.

```{r}
#| label: sim-sd
fit_sd <- function(d) {
  s <- sd(d$y)
  # SE of the sample SD (normal approx) is s / sqrt(2(n-1)).
  list(estimate = s, se = s / sqrt(2 * (nrow(d) - 1)))
}
res_sd <- run_simulation(N_REP, gen, fit_sd, seed = 200)
print_eval(evaluate_estimator(res_sd$estimate, res_sd$se, truth = TRUE_SD))
```

The small negative `bias` on the SD (a fraction of a percent at n = 60) is the
expected finite-sample behaviour — worth knowing, and now demonstrated rather
than taken on faith.

## Takeaways

- The descriptive table needs **no package** — base R is fully sufficient and
  maximally transparent.
- The sample mean is unbiased with correct CI coverage; the sample SD is very
  slightly biased low in small samples, exactly as theory predicts.

```{r}
#| label: session
sessionInfo()
```
